---
title: "R Notebook"
output: html_notebook
---


```{r}
setwd('/Users/gearoidlacey/Documents/College/Masters/Data Analytics')
```

Installing the packages just means that the package is available for use. To actually use it you have to use the *library* command.
The tidyverse package essientally stores all the core packages a data scientist would use in day to day use of R.

```{r}
install.packages('tidyverse')
library('tidyverse')
```

If we want to explicitly access a function of a package use the following syntax

```{r}
ggplot2::ggplot()
```

Its often good practice to visualise the data first. Use the mpg dataset in the ggplot2 package to display a cars engine size in litres vs a cars fuel efficiency. To learn more about the mpg data frame use the command: *?mpg*

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x=displ, y=hwy))
```

The plot roughly shows a *negative relationship* between fuel efficiency and a cars engine size. Therefore as engine size increases, the fuel economy decreases.

Looking at the commands to create the graph; *ggplot()* creates an empty coordinate system (x and y axis with no points or labels) that you can add layers to.

*geom_point()* adds a layer of data points to create a scatter plot. Why use a scatter plot? Because it gives a good initial representation of the data.

Each geom function takes a *mapping argument*. The mapping determines how your data points are mapped to visual properties. The mapping argument is always paired with the aes function. The x and y properties of the aes function determine which values are on the x and y axis.
ggplot will then look for the variables you assign to x and y in the dataframe, in this case called mpg.

Next we will create a graphing template which will act as a general function for creating graphs so we are not replicating code.

```{r}
#ggplot(data = <DATA>) + <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

In the graph of displ vs hwy how do we describe the outliers on the right hand side of the graph.
You can add a third variable into the plots by mapping it to an *aesthetic*. An aesthetic is a visual property of the data points on the graph (size, shape, colour, ect.).

When changing an aesthetic property of a value we say we are *changing the level* of a point.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x=displ, y=hwy, colour=class))
```

By introducing a simple aesthetic such as colouring by class it makes the graph far more informative. So we can see the outliers on the right belong to the 2seater class.

**Note: The colour attribute can be spelt as colour or color**

By adding the aesthetic ggplot will automatically associate a unique colour with each unique class in the dataframe. This process is called *scaling*. ggplot will also add a legend automatically.

Lets change the aesthetic to size.

```{r}
ggplot(data=mpg) + geom_point(mapping = aes(x=displ, y=hwy, size=class))
```

As you can see above we get a warning as mapping an unorded variable (the class of the car) to an ordered aesthetic (size) is not the best approach. As you can see from this graph it is harder to distinguish the classes.

We could also use a transparancey aesthetic (alpha) or the shape aesthetic as you can see below

```{r}
ggplot(data=mpg) + geom_point(mapping = aes(x=displ, y=hwy, alpha = class))

ggplot(data=mpg) + geom_point(mapping = aes(x=displ, y=hwy, shape = class))
```

As you can see a warning message is displayed above the shapes plot as it cannot automatically assign more than **6 unique** shapes so the suv class does not get plotted

The aes function gathers all the aesthetic mappings and passes them to the layers mapping argument.

You can set the aesthetic values manually e.g setting all points to blue

```{r}
ggplot(data=mpg) + geom_point( mapping = aes(x=displ, y= hwy) , colour="blue")
```

As you can see from the code, when you are setting an aesthetic manually it becomes a parameter of the layers function (geom_point) not the aes function. 

Colours are represented as strings, sizes are represented as numbers and shapes are also represented as numbers (each number is associated with a different shape, see *http://r4ds.had.co.nz/data-visualisation.html*)

To see the data types for each column use *str(mpg)*.

One common problem when using ggplot to add a layer is having the plus on a new line. **The plus must be at the end of the line**

```{r}
ggplot(data=mpg) + 
  geom_point(mapping = aes(x=displ, y=hwy, colour = class))

ggplot(data=mpg)
+ geom_point(mapping = aes(x=displ, y=hwy, colour = class))
```

Notice the first graph is created successfully as the plus is in the right place.

Another way to add additional variables apart from aesthetics is to add *facets*. Facets (which are particularly useful for categorical data) splits the plot into multiple plots based on one or many variables. In the python world using matplotlib, these are similar to pair plots.

To do this we use the facet_wrap() function. The first parameter we supply called a formula is *~*. Here, formula is an name of a data structure. This is followed by the variable you want to create subpolts from. The variable should contain discrete data.

Adding a facet is like adding another layer to the plot.

```{r}
ggplot(data=mpg) + geom_point(mapping = aes(x=displ, y=hwy)) + facet_wrap(~ class, nrow = 2)

```

To facet a plot on more than one variable use the facet_grid() function. In this case both variables a separated by the formula operater *~*

```{r}
ggplot(data=mpg) + geom_point(mapping =  aes(x=displ, y=hwy)) + facet_grid(drv ~ cyl)
```

As you can see these graphs give us plenty of information. The labels on the far right refer to the vehicle being 4 wheel drive, front wheel drive or rear wheel drive. The labels along the top represent the number of cylinders a vehicle has.

If you wanted to show your graph without columns you would use the command below.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

```

Again this type of graph enables us to gain a reasonable amount of information from the data points.

A geom is a geometric object used to describe the data in the plot. Different geoms will give different representatinos of the same data.

```{r}
ggplot(data=mpg) + geom_point(mapping=aes(x=displ, y=hwy))

ggplot(data=mpg) + geom_smooth(mapping = aes(x=displ, y=hwy) ,colour="blue")
```

As you can see point graphs are providing a representation of the same data but they are using different geometric objects to do so. The second plot uses the *smooth* geom which fits a smooth line to the data.

Every geom in ggplot takes a mapping argument but not all aesthetics work on all geometric objects. For example you cannot set the shape of a line because that makes no sense as you can imagine.

geom_smooth() has a linetype aesthetic that draws a different type of line for each unique value in the specified column of the dataframe.

```{r}
ggplot(data=mpg) + geom_smooth(mapping = aes(x=displ, y=hwy, linetype=drv, colour= drv))
```

As you can see the plot has provided different lines and colours for each type of vehicle described in the legend.

There are many types of geoms so looking up the official "cheat sheets" is recommended. They can be found at *https://www.rstudio.com/resources/cheatsheets/*.

ggplot will generally group the data when you map an aesthetic to a discrete variable e.g mapping colour to drv. You can do this manually by using the group attribute to draw separate lines for each type of drv.

```{r}
ggplot(data=mpg) + geom_smooth(mapping = aes(x=displ, y=hwy, group = drv))

ggplot(data=mpg) + geom_smooth(mapping = aes(x=displ, y=hwy, group = drv, colour = drv))
```

As you can see, 3 seperate lines are drawn for the 3 types of drv although the group aesthetic does not provide a legend. Therefore it is easier to use aesthetics that perform the mapping automatically as they take care of this for you.

If you add the colour aesthetic as I have done above, a legend is provide and the lines relating to each type of drv are easily distinguished.

You can add multiple geoms to one plot.

```{r}
ggplot(data=mpg) + geom_point(mapping = aes(x=displ, y=hwy, colour=drv)) +
  geom_smooth(mapping = aes(x=displ, y=hwy, colour=drv))
```

As you can see there is code duplication in terms of the format of adding each layer. More specifically the aes function is the same in each geom. To prevent this you can supply the mapping arguments to the ggplot function which will by default apply them to each layer you add.

By adding the mapping argument to the ggplot() function, that mapping will be applied to every layer add to the plot.

```{r}
ggplot(data=mpg, mapping = aes(x=displ, y=hwy, colour=drv)) + geom_point() + geom_smooth()
```

As you can see from the code this is a more concise approach. If you also add a mapping to one of the geoms, that is a local mapping and is applied only to that geom.

```{r}
ggplot(data=mpg, mapping = aes(x=displ, y=hwy, colour=drv)) + geom_point(mapping = aes( shape=drv)) + geom_smooth()
```

As you can see the the shape aesthetics is applied to the geom_point layer only but it is still coloured according to the global attribute. You can also take advantage of this attribute to get the geom layers to display different data.

```{r}
ggplot(data=mpg, mapping = aes(x=displ, y=hwy)) + geom_point(mapping = aes(color=class)) + geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

As you can see, the geom_smooth layer has overwritten the data it displays. It has filtered the dataset to vehicles that are classified as subcompact only and displayed a line based on this data only. Setting se = False means *no confidence interval* is shown.

Many geoms such as bar charts, histograms and boxplots for example provide extra information about your dataset that you can display in your plot. Using the diamonds dataset in the ggplot package we can display a count for each type of diamond even though the counts are not provided with the dataset. This is done by binning.

```{r}
ggplot(data=diamonds) + geom_bar(mapping = aes(x=cut))
```

As you can see from the code the only mapping we provide is that the type of cut should be displayed on the x axis. The y axis generates a frequency count for each type of cut and displays it on the y axis. 

The algorithm used to do this is called a stat (statistical transformation). In this case a count stat is used to create a "mini" dataframe that provides a count for each diamond cut. This new dataset is used to create the bar chart.

It is possible to create the bar chart without a geom by using the statistical transformation only.

```{r}
ggplot(data=diamonds) + stat_count(mapping = aes(x = cut))
```

As you can see emmitting the geom does not matter in this scenario. **Every stat has a default geom and every geom has a default stat**. In this case the default geom for stat_count is the bar geom, therefore a bar chart is displayed.

If, for example, the frequency count was a coloum in the original dataframe we would use the stat ="identity" attribute in the geom function. This will map the bar to the frequency value without having to generate a count.

It is possible to override the default statistical transformation e.g. in terms of a bar chart using a frequency count we can look at the proportion instead.

```{r}
?geom_bar
ggplot(data=diamonds) + geom_bar(mapping = aes(x=cut, y=..prop.., group=1))
```

In this case we need the group attribute to be equal to 1. The reason for this is, by default geom bar will group by the values on the x axis and provide a count of each seperatly. But, as we are looking the proportions we have to look at the proportion of each in relation to the total of all the cuts.

So by setting group = 1 we are really saying that the proportions are calculated based on 1 group (where the group is the total number of cuts), not five as is the case with counts.

A better explanation of this can be found here *https://stackoverflow.com/questions/39878813/r-ggplot-geom-bar-meaning-of-aesgroup-1*.

You can colour a bar chart in two ways using the colour or fill aesthetic. The colour aesthetic will provide a coloured outline whereas the fill aesthetich will colour the entire bar.


```{r}
ggplot(data=diamonds) + geom_bar(mapping = aes(x=cut, colour = cut))

ggplot(data=diamonds) + geom_bar(mapping = aes(x=cut, fill = cut))
```

If we map the fill aesthetic to the variable clarity we get a stacked bar chart.

```{r}
ggplot(data=diamonds) + geom_bar(mapping = aes(x=cut, fill = clarity))
```

In some bars it is hard to differentiate each clarity. To avoid this we can use the position attribute. The position attribute takes 3 values:

*Identity*: overlaps the bars as opposed to stacking on top of each other!

*fill*: similar to stacking but each set of stacked bar charts is the **same height**

*dodge*: provides a seperate bar for each clarity (best option in my opinion)

```{r}
ggplot(data=diamonds) + geom_bar(mapping = aes(x=cut, fill = clarity), position = "Identity")

ggplot(data=diamonds) + geom_bar(mapping = aes(x=cut, fill = clarity), position = "fill")

ggplot(data=diamonds) + geom_bar(mapping = aes(x=cut, fill = clarity), position = "dodge")
```

There is one other position property that is specifically for geom_point. The position *jitter* attribute. 

In the earlier plots using geom_point, each data point in the dataset is rounded so there is some *overplotting* of values. This can make it difficult the see where the majority of the data points lie. To counteract this we use jittering which adds a small amount of random noise to each data point. Therefore the points spread out as they are unlikely to receive the same amount of noise.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x=displ, y=hwy), position = "jitter")
```

As you can see it is easier to see where little clusters of data points are forming.

The default coordinate system used in ggplot2 is the cartesian coordinate system. There are a number of other coordinate systems that have particular uses.

*coord_flip()*: switches the x and y axis. Useful if you want to have horizontal box plots

```{r}
g <- ggplot(data=mpg, mapping = aes(x=class, y=hwy))  + geom_boxplot()
g

g + coord_flip()
```

*coord_quickmap()*: sets the aspect ratio for spatial data.This is not covered in the book.

*coord_polar()*: uses polar coordinates which reveal an interesting connection between a bar chart and a Coxcomb chart.

```{r}
?theme
bar <- ggplot(data=diamonds) + geom_bar(mapping = aes(x=cut, fill=cut)) + labs(x=NULL, y=NULL)
bar + coord_flip()
bar + coord_polar()
```

Based on all the information we have gained so far a new graph template would consist of seven parameters as follows:

**Data**
**Geom function**
**Mappings**
**Stat**
**Position**
**Facet function**
**Coordinate system**

```{r}
#ggplot(data=<DATA>) + <GEOM_FUNCTION>(mapping=aes(<MAPPINGS>), stat=<STAT>, position=<POSITION>) + <COORDINATE_SYSTEM> + <FACET_FUNCTION>
```



```{r}

```
