---
title: "R Notebook"
output: html_notebook
---

```{r}
setwd('/Users/gearoidlacey/Documents/College/Masters/Data Analytics')

library(tidyverse)
```

Exploratory Data Analysis (EDA) is an iterative process:
  1. Generate a research question.
  2. Visualize and transform the data to gain insights.
  3. Use the new information to re-examine the data.
  
EDA helps you understand the data. There are two main questions you shoud ask about the data.
  1. What type of variation occurs within the variables
  2. What type of covariation occurs within the variables
  
  
Variation is the tendency of the values of **a variable** to change from measurement to measurement.
Each variable has its own pattern of variation which is often best described using a plot.

Visualising the distribution of a variable depends whether its numerical or categorical. A categorical variable (e.g. of data type factor in R) has a finite set of values it can be. A bar chart is best used to describe categorical variables.


```{r}
ggplot(data = diamonds) + geom_bar(mapping = aes(x=cut))
```

The height of each bar represents the number of diamonds with the specified cut. This can be done using the count function in the dplyr package.

```{r}
diamonds %>%
  dplyr::count(cut)
```

This shows the information the graph is based on. A variable is continuous (normally a numerical value) if theres an infinite amount of values the variable can be. *Histograms* are often used to look at the distribution of a continuous variable.

```{r}
ggplot(data = diamonds) + geom_histogram(mapping = aes(x=carat), binwidth = 0.5)
```

Again this information can be got manually.

```{r}
diamonds %>%
  count(cut_width(carat, 0.5))
```

This information shows us that there was one diamond with a carat value between 4.25 and 4.75. The *cut_width* function basically converts the continuous variable to a discrete variable by breaking the carat values into ranges that increase in .5 increments. 

Therefore the first range starts at -.25 up to .25 carats. The next range is .25 carats up to .75 carats and so on.

A histogram divides the x axis into evenly space bins then uses the height of the bar to diplay information. You should always experiment with the value of the *bin_width* parameter as the height of the bars will depend on this.

```{r}
ggplot(data = diamonds) + geom_histogram(mapping = aes(x = carat), binwidth = 0.1)
```

As the bin_width has changed, so has the height of the bars. The *default number of bins is 30*, but this is overwritten when the bin_width is changed.

If you want multiple histograms in the same plot use *geom_freqpoly*.

```{r}
ggplot(data = diamonds) + geom_freqpoly(mapping = aes(x = carat, colour=cut), binwidth = .4)
```

A crucial part of EDA is to ask questions of the data.

```{r}
ggplot(data = diamonds) + geom_histogram(mapping = aes(x = carat), binwidth = .01)
```

Some questions you could ask of this graph is: Why are diamonds with a whole number carat more important. Why are there no diamonds greater than 3 carats. Why is there a high frequency of diamonds after each of the major peaks.

Clusters of similar values generally suggest there are subgroups in the data.

Outliers are data points that dont fit the general pattern. They are not always data entry errors which might be most peoples initial assumption. In my opinion scatterplots are a good way to find outliers.

Sometimes when the counts on the y axis are so high the smaller bars cannot be seen. To change this you can "zoom in" on the values using the *coord_cartesian()* by setting a limit on the y axis.

```{r}
ggplot(data = diamonds) + geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0,50))
```

ggplot also has *xlim* and *ylim* attributs but the work different in that they remove the values outside the limits as opposed to just zooming in.

As we can see from the graph there are 3 unusual values, 0, approximately 30 and approximately 60. We can now filter to look at these values specifically

```{r}
?diamonds

diamonds %>%
  filter(y == 0 | y > 20) %>%
  arrange(desc(y))
```

From this we can see that the two diamonds that were above 30 in the y variable are considered ideal and premium. Diamonds cannot have a width of 0 mm so those values are incorrect. A y value of 58mm is nearly 2 inches long which would seem highly unusual.

To combat missing values its generally best to keep the data all together for as long as possible and treat removing values as the last option. A good option to avoid removing values is to use to mutate function to copy to "bad" data into a new column but replace the bad values with NA using the *ifelse* function.

```{r}
(diamonds2 <- diamonds %>%
  mutate(y = ifelse( y < 3 | y > 20, NA , y)))
```

The ifelse function reads like so:
  If y is less than 3 or greater than 20 assign it the value NA, otherwise leave the value as it is.
  
Although ggplot will remove missing values before plotting, it will alert you to the fact that it has done so.

```{r}
ggplot(data = diamonds2) + geom_point(mapping = aes(x=x, y=y))
```

To stop that warning, add *na.rm = TRUE* in the geom layer.

Sometimes we will want to compare missing values with non-missing values. For example comparing the scheduled departure time for cancelled and uncancelled flights.

```{r}
flights %>%
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60) %>%
  
  ggplot(mapping = aes(sched_dep_time)) + 
    geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4)
  
```

As there are more flights that have departed than have been cancelled, this graph is not very informative.

As we have observed, variation is how the values of one specific variable change from measurement to measurement. But now we want to look at **covariation** which is **how the values of more than 1 variable** change from measurement to measurement in a related way.

The best way to determine covariation between 2 or more variables is, as you may have guessed, to visualise the data. Observing the relationship between a categorical and continuous variable.

We could use the freqpoly geom but its harder to distinguish the smaller groups.

```{r}
ggplot(data = diamonds) + geom_freqpoly(mapping = aes(x = price, colour = cut), binwidth = 500)
```

As the counts differ so much hard to see if any covariation exists. Therefore we need to change the y axis to density.

```{r}
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```

This graph suggests fair diamonds have the highest average price. A better way to observe a continuous variable vs a categorical variable would be to use a boxplot. This particular section is explained far better in the book so I advise you to look at section 7.5.1.

The boxplot is often used as it provides so much information in one plot i.e the interquartile range and outliers etc.

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = price, colour = cut)) + geom_boxplot()
```

This plot also says that better quality diamonds are cheaper which is intresting? The points above each of the plots are 1.5 times away from the IQR. These points are considered unusual so they are plotted individually. The top end of each of the lines protruding from the plots indicates where the last non-outlier point.

As you can see the categorical variables are in order but this wont always be the case. If this is not the case use the *reorder()* function.

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + geom_boxplot()

?reorder
ggplot(data = mpg, mapping = aes(x = reorder(class, hwy, median), y = hwy)) + geom_boxplot()
```

Reorder works by specifying the variable you want to reorder and how you want to reorder it by. In our case we want to reorder the classes based on the median hwy values.

If the labels on the x axis were very long we could use the coord_flip() function

```{r}
ggplot(data = mpg, mapping = aes(x = reorder(class, hwy, median), y = hwy)) + geom_boxplot() + coord_flip()
```

Next we will look at covariation between to categorical variables. So if theres categorical variables on the x and y axis we need to get a count for each observation.

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = color)) + geom_count()
```

This could also be visualised using a tile graph.

```{r}
diamonds %>%
  count(color,cut) %>%
  ggplot(mapping = aes(x = color, y = cut)) + geom_tile(mapping = aes(fill = n))
```

When looking at two continuous variables the natural approach is to use a scatterplot.

```{r} 
?geom_point

ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
```

In terms of looking at specific values this plot is not very informative. One way to resolve this is to change the value of alpha.

```{r}
?geom_point

ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price), alpha = 1/100)
```

An alternative solution would be to use bins (1 Dimensionally) which we have done with histograms and freqpoly geoms. We can also bin in 2D using *geom_bin2d()* and *geom_hex()*. geom_bin2d() creates rectangular bins, geom_hex() creates hexagonal bins.

To use geom_hex() you need to install the *hexbin* package.

```{r}
ggplot(data = diamonds) + geom_bin2d(mapping = aes(x = carat, y = price))

#install.packages('hexbin')
library('hexbin')

ggplot(data = diamonds) + geom_hex(mapping = aes(x = carat, y = price))
```

Another option is to bin one of the continuous variables so it is treated as a categorical variable using the cut_width() function.
```{r}
?cut_width
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) + geom_boxplot(mapping = aes( group = cut_width(carat, width = 0.1)))
```

Some of the boxes are approximately the same width so its difficult to tell that they summarise a different number of points. We can counteract this by setting *varwidth = TRUE* which makes the width of the boxplot proportional to the number of points **OR** using the *cut_number()* function which displays approximately the same number of points in each bin.

```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes( group = cut_number(carat, 20)))
```

Patterns provide useful information for data scientists as they reveal covariation. Models are used to extract patterns from data. We will come back to this in more detail later.

Over time it almost becomes instinctive what the main paramters function take are. For ggplot() the two main ones are *data* and *mapping*. The main parameters for the aes() are *x* and *y*. We will no longer supply these parameter names.

Chapter 8 of the book is more theoretical so I advise you to read the book itself for that chapter,  *http://r4ds.had.co.nz/workflow-projects.html*. Noticeably when setting paths in R you can use both foward or backslashes in the path but you have to use double backslashes as single backslashes have a special meaning.
