---
title: "R Notebook"
output: html_notebook
---

Relational data: multiple tables of data.

Relations are defined between a pair of tables.

There are 3 types of joins when working with relational data:
    1. Mutating joins: add new variables to one data frame from matching observations in another.
    2. Filtering joins: filter observations from 1 data frame based on whether or not they match observations in another.
    3. Set operations: treat observations as if they are set elements
    
    
```{r}
library(tidyverse)
library(nycflights13)
```

The nycflights package has four tables. Section 13.2 shows how the tables relate to each other.

```{r}
airlines
airports
planes
weather

```

The variables used to connect each table are called keys. Keys are used to uniquely identify observations. There are 2 types of keys:
    1. Primary key: used to uniquely identify an observation in its own table.
    2. Foreign key: used to uniquely identify observations in other tables.
    
When you've identified the primary keys in each table, verify it by taking a count of the primary keys.

```{r}
planes %>%
  count(tailnum) %>%
  filter(n > 1)

weather %>%
  count(year,month, day ,hour, origin) %>%
  filter(n > 1)

airports %>%
  count(faa) %>%
  filter(n > 1)

airlines %>%
  count(carrier) %>%
  filter(n > 1)
```

If a table does not have a primary key you can add one using *mutate()* and *row_number()*

```{r}
?row_number
```

Mutating joins allow us add variables from one table to another where the observation keys match.
Create a smaller dataset to see how this works.

```{r}
flights2 <- flights %>% 
  select(year:day, hour, origin, dest, tailnum, carrier)
flights2
```

Lets combine the airlines and flights2 dataframe so flights2 contains the airline name. This can be done 2 ways. The select statement removes the origin and dest columns from the dataframe.

```{r}
flights2 %>%
  select(-origin, -dest) %>%
  left_join(airlines, by = 'carrier')

?dplyr::mutate
flights2 %>%
  select(-origin, -dest) %>%
  mutate(name = airlines$name[match(carrier, airlines$carrier)])
```

There are multiple types of mutating joins. If you are familiar with joins in SQL databases this section will be relatively straigthforward.

Inner-join: matches observations where the keys are equal. Therefore unmatched joins are not included in the result.

Outer-join: keeps observations that appear in **at least 1 of the tables.** There are 3 types of outer joins:
    1. Left join: keeps all observations in x and observations that match in y.
    2. Right join: keeps all observations in y and observations that match in x.
    3. Full join: keeps all observations in x and y.

```{r}
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     3, "x3"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     4, "y3"
)

x %>%
  inner_join(y, by ='key')

x %>%
  right_join(y, by ='key')

x %>%
  left_join(y, by ='key')

x %>%
  full_join(y, by ='key')
```

I encourage you to look at the graphic the authors provide in section 13.4 of the book.

**Note when you join on duplicate keys in two tables you key all possible results of that join.**

You dont have to specify what column you want to join by. If you dont specify, R will join on all columns that match between both tables as illustrated below

```{r}
flights %>%
  left_join(weather)


flights %>%
  left_join(airports, by = c('dest' = 'faa'))
```

The second join works by joining on the dest and faa columns but only the dest column is displayed in the result.

The *merge()* function in the base package does the same thing as the aforementioned joins.

Filtering joins match observations the same way as mutating joins but affect the observations not the variables. There are two types of filtering joins:

    1.Semi joins: **keep** all the observations in x that have a matching key in y
    2.Anti join: **drops** all the observations in x that have a matching key in y
    
I suggest looking at the book for this section. Lets move onto set operations which are self explained.

```{r}
df1 <- tribble(
  ~x, ~y,
   1,  1,
   2,  1
)
df2 <- tribble(
  ~x, ~y,
   1,  1,
   1,  2
)

intersect(df1,df2)

union(df1,df2)

setdiff(df1,df2)

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```


