---
title: "R Notebook"
output: html_notebook
---

To work with dates use the *lubridate* package.

```{r}
library(tidyverse)
library(lubridate)
library(nycflights13)
```

To get the current date use the *today()* or *now()* functions.

```{r}
today()

now()
```

You can use the following function to create a date from a string in a specific format.

```{r}
ymd('2018 February 16')

mdy('February 16 2018')

dmy('16 February 2018')

ymd('20180216')
```

To create a date-time add _hms to each function name. You can also specify the timezone by using the tz parameter.

```{r}
ymd_hms('2018-02-16 16:58:00')

ymd_hm('2018-02-16 16:58', tz='GMT')
```

If the date/datetime is split into different parts use the *make_date()* or *make_datetime()* functions.

```{r}
flights %>%
  select(year, month, day, hour, minute) %>%
  mutate(departure = make_datetime(year, month, day, hour, minute))
```

The following two code blocks are taken directly from the book where they look at the distribution of departure times.

```{r}
make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights %>% 
  filter(!is.na(dep_time), !is.na(arr_time)) %>% 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>% 
  select(origin, dest, ends_with("delay"), ends_with("time"))
```



```{r}
flights_dt %>% 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 86400) # 86400 seconds = 1 day
```

For the rest of this section I would encourage you the read section 16.3 onwards as its more applicable if your doing time series analysis in my opinion.

```{r}
```

```{r}

```

