---
title: "R Notebook"
output: html_notebook
---

As previously mentioned the pipe operater *%>%* helps us to stop having to assign the results of each operation to a new variable.

The pipe operator is located in the *magrittr* package which is loaded by default in most of the main packages.

```{r}
library(magrittr)
library(nycflights13)

grouped_flights <- group_by(flights,carrier)

count(grouped_flights, carrier)

flights %>%
  group_by(carrier) %>%
  count(carrier)
```

Instead of having to assign the result of the grouping to a new object we can use the pipe operator to continue with our analysis. In the background magrittr basically makes the code we supply into function calls

It is worth noting that when you make a data frame from another data frame R is efficient in terms of memory as it really only stores the different columns between the two.

```{r}
#install.packages('pryr')
library(pryr)
?object.size
format(object.size(x = flights), units = "auto")
format(object.size(x = grouped_flights), units = "auto")
pryr::object_size(grouped_flights, flights)
```

As you can see the combined size of the data sets is 42MB. 

Getting back to the pipe operator, there are numerous cases where you should **not** use a pipe.

    1. When there are more than 10 steps. Makes debugging difficult if theres more than 10 steps.
    2. If you have multiple inputs/outputs.
    3. Pipes are linear, expressing complex relationships with them will normally give confusing results.
    
If your using a pipe and you want to see the result of one of the transformations in you code use *%T>%*.

```{r}
library(forcats)

gss_cat %>%
  mutate(relig = fct_lump(relig, n=5)) %>%
  group_by(relig) %>%
  count() %T>%
  plot()
```

As you can see we now get the counts for each religion as well as the plot.

If your working with a data structure that does not have a data frame based API e.g. referring to variables in a list, use  * %$% *.

```{r}
?`%$%`

typeof(mtcars)

mtcars %$%
  cor(disp, mpg)
```

