---
title: "R Notebook"
output: html_notebook
---

To work with strings we will need to use the *stringr* library

```{r}
library(tidyverse)
library(stringr)
```

You can wrap strings in double or single quotes (" or '). As I primarily use python, I tend to use single quotes.

If you forget to close a quote you will see a plus sign on the left which is generally the case when R expects something its not getting. If this happens press the escape key, fix your code and run it again.

```{r}
(string1 <- "uses double quotes")
(string2 <- 'uses single quotes')
```

To include a " or a ' in the string put a backslash before it. (Appologies for the incorrect gramatical example!)

```{r}
(string3 <- 'uses single quote\'s')
```

The printed representation of a string is not the same as the raw representation of a string as the printed representation shows the escape characters.

```{r}
string4 <- '\''
string4

writeLines(string4)
```

As you can see the first representation shows the string wrapped in quotes whereas the second representation only shows the string itself.

Base R has string functions but they are inconsistent so use stringr. All stringr functions start with *str_*.

Two useful functions when working with strings: getting the string length (*str_length*) and concatenating strings (*str_c*).

```{r}
str_length('This is R programming')

str_c('Hello', 'world', sep =' ')
```

You can collapse a vector of strings into one string.

```{r}
str_c('H', 'e', 'l', 'l', 'o', collapse = TRUE)
```

To get a substring use the *str_sub()* function. You specify the string you want to substring as well as the start and end positions of the substring

```{r}
string5 <- 'This is R programming'

str_sub(string5, 11,17)
```

Regular expressions allow you define patterns in strings. Lets match a parameter in a string using the *str_view()* function.

```{r}
strings <- str_split(string5, pattern = ' ')
str_view(strings, "ram")
```

Regular expressions can be hard to get your head around so I suggest reading the book for the extensive section it has on it.

```{r}

```



```{r}

```



```{r}

```



```{r}

```


